---
title: "Lab 5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 8: Hypothesis Testing

A hypothesis is a claim about a property of a population. A hypothesis test is a standard procedure for testing a claim about a postulation. It is also called a test of significance.

## 8-2: Testing a Claim About a Proportion.

The test statistic is 

\[
z = \frac{\hat{p} - p}{\sqrt{\frac{pq}{n}}}
\]

Where:
- \( \hat{p} \): Sample proportion.
- \( p \): Population proportion.
- \( q = 1 - p \): Complement of the population proportion.
- \( n \): Sample size.

The following requirements must be met:
1. The data come from a simple random sample.
2. The conditions for a binomial distribution are satisfied. 
3. np≥5  and nq≥5  

The R command for testing the hypothesis about a population proportion is 

<span style="color:red">prop.test(x, n, p, correct =, alternative = )</span>

Note:
- correct = TRUE uses a continuity correction factor
- correct = FALSE does not use a continuity correction factor
- alternate = "two.sided" (default)  or “greater” or “less”

**Example 1**

A study of sleepwalking or “nocturnal wandering” was described in Neurology magazine, and it included information that 29.2% of 19,136 American adults have sleepwalked. Would a reporter be justified in stating that “fewer than 30% of adults have sleepwalked”? Let’s use a 0.05 significance level to test the claim that for the adult population, the proportion of those who have sleepwalked is less than 0.30.

**Solution**

```{r}
prop.test(0.292*19136, 19136, p = 0.3, alternative = "less" )
```
## 8-3: Testing Claims About a Mean when the population standard deviation is unknown

The Test Statistic is 

\[
t = \frac{\bar{x} - \mu}{\frac{s}{\sqrt{n}}}
\]

Where:
- \( \bar{x} \): Sample mean.
- \( \mu \): Population mean.
- \( s \): Sample standard deviation.
- \( n \): Sample size.

Requirements

1.	The sample is a simple random sample
2.	The population is normally distributed or n > 30

The R command for testing the hypothesis about the mean when the population standard deviation is unknown is 

<span style="color:red">t.test(x, mu = , alternative = )</span>

Note:
•	alternate = "two.sided" (default)  or “greater” or “less”

**Example 1**

Listed below are amounts of sleep (hours) for randomly selected adult subjects included in the National Health and Nutrition Examination Study. Here are the unrounded statistics for this sample: n = 12,   x = 6.83333333 hours, s = 1.99240984 hours. A common recommendation is that adults should sleep between 7 hours and 9 hours each night. Use the p-value method with a 0.05 significance level to test the claim that the mean amount of sleep for adults is less than 7 hours.

- 4   8   4   4   8   6   9   7   7   10   7   8

**Solution**

```{r}
data1 <- c(4, 8, 4, 4, 8, 6, 9, 7, 7, 10, 7, 8)
 
t.test(data1, mu = 7, alternative = "less")
```
## Section 8.4: Testing a Claim About a Standard Deviation or Variance

Test Statistic:

\[
\chi^2 = \frac{(n-1) s^2}{\sigma^2}
\]

Where:
- \( \chi^2 \): Chi-square statistic.
- \( n \): Sample size.
- \( s^2 \): Sample variance.
- \( \sigma^2 \): Population variance.

The R command for the Chi Square statistic is given by

<span style="color:red">Chi_sq <- ((n - 1) * sd_x^2)/sigma_x^2

where sd_x and sigma_x are the mean and standard deviation respectively.
The R command for the p-value associated with the Chi square statistic is given by

<span style="color:red">pchisq(Chi_sq, n-1, lower.tail = TRUE)

**Example 1**

Listed below are pulse rates in beats per minute (bpm) for a simple random sample of females from Data Set 1 “Body Data” in Appendix B. Use a 0.05 significance level to test the claim that females have pulse rates with a standard deviation that is less than the value σ=11.3 bpm from the population of males.

**Solution**
```{r}
Data1 <- c(76, 76, 86, 74, 66, 62, 78, 68, 62, 62, 74, 82, 54, 74, 74, 84, 60, 52, 84, 66, 56, 66, 84, 64)
n <- length(Data1)
sd_x <- sd(Data1)
sigma_x <- 11.3
Chi_sq <- ((n - 1) * sd_x^2)/sigma_x^2
Chi_sq

p_value <- pchisq(Chi_sq, n-1, lower.tail = TRUE)
p_value
```
**Example 2**

Repeat the hypothesis test in Example 1 by constructing a suitable confidence interval.

**Solution**
```{r}
chisq_L <- qchisq(0.05, n-1, lower.tail = TRUE)
chisq_L

chisq_R <- qchisq(0.05, n-1, lower.tail = FALSE)
chisq_R

Lower <- sqrt((n-1)*sd_x^2/chisq_R)
Lower

Upper <- sqrt((n-1)*sd_x^2/chisq_L)
Upper
```

## Summary of R Commands
### <span style="color:red">R command</span>	Task

- <span style="color:red">prop.test(x, n, p, correct =, 
alternative = )</span>
	Tests the hypothesis about a population proportion
- <span style="color:red">t.test(x, mu = , alternative = )</span>
	Tests the hypothesis about the mean when the population standard deviation is unknown
- <span style="color:red">Chi_sq <- ((n - 1) * sd_x^2)/sigma_x^2</span>
	Computes the chi square statistic
- <span style="color:red">pchisq(Chi_sq, n-1, lower.tail = TRUE)</span>
	Computes the p-value associated with the Chi square statistic







