---
title: "Lab 9"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 12: Analysis of Variance

Analysis of Variance presents methods for testing equality of three or more population means.

## 12-1: One-Way ANOVA

One-Way Analysis of Variance is a method for testing the equality of three or more population means by analyzing the sample variances.  We test the following hypotheses:

**Null and Alternative Hypotheses:**

- \( H_0: \mu_1 = \mu_2 = \mu_3 = \cdots = \mu_k \)
- \( H_a: \) At least one pair of means are not the same.

To perform the test, the raw data must be organized in the right format. For example, if we have the data in the form:

| Column1 | Column2 | Column3 | Column4 |
|---------|---------|---------|---------|
| \( a_1 \) | \( b_1 \) | \( c_1 \) | \( d_1 \) |
| \( a_2 \) | \( b_2 \) | \( c_2 \) | \( d_2 \) |
| \( a_3 \) | \( b_3 \) | \( c_3 \) | \( d_3 \) |
| \( a_4 \) | \( b_4 \) | \( c_4 \) | \( d_4 \) |
| \( a_5 \) | \( b_5 \) | \( c_5 \) | \( d_5 \) |

Then we write the following R code:

<span style="color:red">Column1 <- c(a_1,a_2,a_3,a_4,a_5)</span>

<span style="color:red">Column2 <- c(b_1,b_2,b_3,b_4,b_5)</span>

<span style="color:red">Column3 <- c(c_1,c_2,c_3,c_4,c_5)</span>

<span style="color:red">Column4 <- c(d_1,d_2,d_3,d_4,d_5)</span>

<span style="color:red">Data1 <- c(Column1, Column2, Column3, Column4)</span>

<span style="color:red">Column_names <- c(rep("Column1", 4), rep("Column2", 4), rep("Column3", 4), rep("Column4", 4))</span>

<span style="color:red">Data <- data.frame(Data1, Column_names)</span>

Please note that the last line of the code above is only for the display of the data. It is totally optional, and can be left out without affecting the results.

We can obtain a boxplot of the data to get a visual picture of how the populations compare. 

Next, we perform Levene’s Test for homogeneity of variances by first installing the “car”package:

<span style="color:red">install.packages (“car”)</span>

and typing the following code:

<span style="color:red">LeveneTest(Data1, Column_names)

A nonsignificant result is required before we can proceed to perform the ANOVA test.  We preform the ANOVA test using the following R command:

<span style="color:red">model <- aov(Data1 ~ Column_names)</span>

To obtain the output, we type the following command:

<span style="color:red">summary(model)</span>

If the output shows significance, we can perform Tukey’s HSD Test by typing the following command:

<span style="color:red">TukeyHSD(model)</span>

**Example 1**
Use the head injury criterion (HIC) measurements listed in the table below and use a significance level of 0.05 to test the claim that the four samples come from populations with means that are all equal.

| Small | Midsize | Large | SUV |
|-------|---------|-------|-----|
| 253   | 117     | 249   | 121 |
| 143   | 121     | 90    | 112 |
| 124   | 204     | 178   | 261 |
| 301   | 195     | 114   | 145 |
| 422   | 186     | 183   | 198 |
| 324   | 178     | 87    | 193 |
| 258   | 157     | 180   | 193 |
| 271   | 203     | 103   | 111 |
| 467   | 132     | 154   | 276 |
| 298   | 212     | 129   | 156 |
| 315   | 229     | 266   | 213 |
| 304   | 235     | 338   | 143 |


**Solution**
```{r}
small <- c(253, 143, 124, 301, 422, 324, 258, 271, 467, 298, 315, 304)
midsize <- c(117, 121, 204, 195, 186, 178, 157, 203, 132, 212, 229, 235)
large <- c(249, 90, 178, 114, 183, 87, 180, 103, 154, 129, 266, 338)
SUV  <- c(121, 112, 261, 145, 198, 193, 193, 111, 276, 156, 213, 143)
head_injury <- c(small, midsize, large, SUV)
car_model <- c(rep("small", 12), rep("midsize", 12), rep("large", 12), rep("SUV", 12))
HIC <- data.frame(head_injury, car_model)
head_injury <- c(small, midsize, large, SUV)
car_model <- c(rep("small", 12), rep("midsize", 12), rep("large", 12), rep("SUV", 12))

boxplot(head_injury ~ car_model)

library(car)
leveneTest(head_injury, car_model)

ANOVA1 <- aov(head_injury ~ car_model)

summary(ANOVA1)

TukeyHSD(ANOVA1)

```

## 12-2: Two-Way ANOVA

A two-way Analysis of Variance is used with data that are partitioned into categories according to two factors. First, we test for the interaction between the two factors: the row and the column. We then test for an effect from the row factor and the column factor. 
We first obtain the interaction plot by typing the following command:

<span style="color:red">interaction.plot(variable_name, row_name, column_name)</span>

We perform the two-way ANOVA test by typing the following command

<span style="color:red">model <- aov(variable_name ~ row_name + column_name + row_name * column_name)</span>

As in section 12-1, we can obtain the output by typing the following command:

<span style="color:red">summary(model)</span>

**Example 1**

Given the Table 12-3 measurements of crash forces on the left and right femurs in car crash tests, use two-way analysis of variance to test for an interaction effect, an effect from the row factor of femur side (left, right), and an effect from the column factor of vehicle size category (small, midsize, large, SUV). Use a 0.05 significance level.

**Solution**

```{r}
Femur_left <- c(1.6, 1.4, 0.5, 0.2, 0.4, 0.4, 0.7, 1.1, 0.7, 0.5, 0.6, 1.8, 0.3, 1.3, 1.1, 0.4, 0.4, 0.6, 0.2, 0.2)
Femur_right <- c(2.8, 1.0, 0.3, 0.3, 0.2, 0.6, 0.8, 1.3, 0.5, 1.1, 1.5, 1.7, 0.2, 0.6, 0.9, 0.7, 0.7, 3.0, 0.2, 0.2)
Femur <- c(Femur_left, Femur_right)
Femur_side <- c(rep("Left", 20), rep("Right", 20))
Vehicle_Category <- c(rep("Small", 5), rep("Midsize", 5), rep("Large", 5), rep("SUV", 5),rep("Small", 5), rep("Midsize", 5), rep("Large", 5), rep("SUV", 5))
Crash_Test_Force <- data.frame(Femur, Femur_side, Vehicle_Category)

interaction.plot(Femur_side, Vehicle_Category, Femur)

ANOVA1 <- aov(Femur ~ Femur_side + Vehicle_Category + Femur_side * Vehicle_Category)

summary(ANOVA1)
                                         
```
## Summary of R Commands
### <span style="color:red">R command</span>	Task

- <span style="color:red">
install.packages (“car”)</span>
	Installs the “car” package that is needed to perform Levene’s test.
- <span style="color:red">LeveneTest(Data1, Column_names)</span>
	Performs Levene’s Test for the homogeneity of variances.
- <span style="color:red">model <- aov(Data1 ~ Column_names)</span>
	Performs the One-Way ANOVA test
- <span style="color:red">TukeyHSD(model)</span>	Performs Tukeys HSD test to identify which populations are significantly different.
- <span style="color:red">interaction.plot(variable_name, row_name, column_name)</span>
	Provides the interaction plot for a two-way ANOVA
- <span style="color:red">model <- aov(variable_name ~ row_name + column_name + row_name * column_name)</span>
	Performs the two-way ANOVA test
- <span style="color:red">summary(model)</span>	Gives the output of the One-Way ANOVA test



